---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sigsolve

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/sigsolve)](https://CRAN.R-project.org/package=sigsolve)
[![R-CMD-check](https://github.com/selkamand/sigsolve/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/selkamand/sigsolve/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/selkamand/sigsolve/branch/main/graph/badge.svg)](https://app.codecov.io/gh/selkamand/sigsolve?branch=main)
![GitHub Issues or Pull Requests](https://img.shields.io/github/issues-closed/selkamand/sigsolve)
[![code size](https://img.shields.io/github/languages/code-size/selkamand/sigsolve.svg)](https://github.com/selkamand/sigsolve)
![GitHub last commit](https://img.shields.io/github/last-commit/selkamand/sigsolve)
<!-- badges: end -->

> [!WARNING]  
> This package is in early development and not yet ready for use

The **sigsolve** R package implements common mutational signature fitting 
algorithms that allow you to identify the most likely combination of known mutational signatures 
which explain the mutational profile observed in a sample of interest.

> [!TIP]  
> sigsolve is exclusively a signature fitting tool. For turnkey end-to-end signature analysis 
> we recommend [sigscreen](https://github.com/selkamand/sigscreen)

Unlike most other implementations, sigsolve expects no specific signature collection or feature space. 
Users supply a catalogue of counts & a matching signature collection, so any type of signature analysis / feature space is supported.

## Installation

You can install the development version of sigsolve like so:

``` r
if (!require("remotes"))
    install.packages("remotes")

remotes::install_github("selkamand/sigsolve")
```

## Quick Start

There are 2 inputs required. 

1. **observed mutation counts**: A named numeric vector where names represent channels and values represent counts
2. **signature collection**: A channel X signature matrix where values represent fractions.


```{r}
library(sigsolve)

catalogue <- c(channel1 = 20, channel2 = 104, channel3 = 10)

signatures <- matrix(
  c(0, 0.9, 0.1, 0.98, 0.01, 0.01, 0.25, 0.5, 0.25),
  nrow = 3L,
  ncol = 3L,
  dimnames = list(c("channel1", "channel2", "channel3"), c("Sig1", "Sig2", "Sig3"))
)

sig_fit(catalogue, signatures, method = "nnls")
```
